---
title: "Expectation violation review"
subtitle: "Descriptive analyses"
author: "José C. García Alanis"
date:  "`r Sys.Date()`"
output: html_document
---

Created with `r R.version.string`

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)

# set working directory
if(interactive()){
  path <- dirname(rstudioapi::getSourceEditorContext()$path)
} else {
  path <- normalizePath('./')
}
setwd(path)

```


# Seu up environment

```{r, message = FALSE}
# load necessary packages
require(dplyr)
require(bibliometrix)

# options for plotting
# windowsFonts(Mukta=windowsFont("Mukta"))
require(ggplot2)
font_fam <- 'Mukta'

# get other packages for plotting
library(viridis)
library(grid)
library(gridExtra)
library(ggrepel)

```


# Load the data

```{r}
path_to_data <- './data/data-2021-01-04_filtered.RData'

# imports data frame with all identified publications
load(path_to_data)

# create results dir
if (!dir.exists('./results')) {
  dir.create('./results')
  dir.create('./results/figures')
}
```

# Compute results to generate Figure 1

Compute the annual scientific production (i.e., number of publications per year) 
since 1990

```{r}
ann_prod <- M %>% 
  # use PY (Year Published) WoS field tag
  group_by(PY) %>%
  summarise(n_articles = sum(!is.na(PY)))
```

## Generate Figure 1

```{r fig.align = 'center'}
ann_prod_plot <- ggplot(ann_prod, aes(x = PY, y = n_articles)) + 
  geom_point(alpha = 0.85, size = 1.5) +
  # geom_smooth(method = lm, formula = y ~ splines::ns(x, 3), 
  #             size = 0.75, alpha = 0.25, color = '#D64B40FF') +
  geom_line(size = 0.75) +
  geom_segment(aes(x = -Inf, y = 0, xend = -Inf, yend = 150),
               color = 'black', linetype = 1) +
  geom_segment(aes(x = 1990, y = -Inf, xend = 2020, yend = -Inf),
               color = 'black', linetype = 1, ) +
  scale_x_continuous(breaks = c(seq(1990, 2020, 6))) +
  scale_y_continuous(breaks = c(seq(0, 150, 25))) +
  theme(plot.margin = unit(c(5, 30, 5, 5), 'pt'),
        panel.background = element_blank(),
        plot.title = element_text(color = 'black', size = 18, face = 'bold', 
                                  hjust = 0, family=font_fam),
        axis.title.x = element_text(color = 'black', size = 16, face = 'bold',
                                    margin = margin(t = 10), family=font_fam),
        axis.title.y = element_text(color = 'black', size = 16, face = 'bold',
                                    margin = margin(r = 10), family=font_fam),
        axis.text.x = element_text(color = 'black', size = 14,
                                   family=font_fam),
        axis.text.y = element_text(color = 'black', size = 14, family=font_fam),
        panel.grid.major.x = element_line(size = 0.5, linetype = 'solid',
                                          colour = "gray98"),
        panel.grid.major.y = element_line(size = 0.5, linetype = 'solid',
                                          colour = "gray98"),
        panel.grid.minor.y = element_line(size = 0.25, linetype = 'solid',
                                          colour = "gray98"),
        legend.title = element_blank()) +
  labs(y = 'Number of publications', x = 'Year', 
       title = 'Annual scientific production'); ann_prod_plot
ggsave(filename = './results/figures/Fig_1.tiff',
       plot = ann_prod_plot, 
       width = 16, height = 10, units = 'cm',
       dpi = 1200)
```


### Follow up anaylses

Get statistics to describe the increase in publications since 1990

```{r}
# fit cubic growth model
model_01 <- lm(n_articles ~ splines::ns(PY, df = 3),
                data = ann_prod)
# get F-statistic
anova(model_01)

# create data frame to store predictions
N <- length(ann_prod$PY)
x <- ann_prod[which.max(complete.cases(ann_prod)), ]
df <- do.call(rbind, lapply(1:N, function(i) x))
df$PY <- ann_prod$PY

# compute spline effect
df$art.hat <- predict(model_01, newdata = df)
```

Plot results

```{r fig.align='center'}
with(df, plot(PY, art.hat, type = "l", lwd = 2, 
              ylab = "", xlab="Year", 
              main = "Relative spline term"))
mtext("Spline contribution\nto the link function", side=2, line=2)

# plot estimated change per year
dYear <- diff(df$PY[1:2])
delta <- diff(df$art.hat)/dYear
year <- (df$PY[-N] + df$PY[-1]) / 2
plot(year, delta, type="l", lwd=2, 
     ylab="Change per year", xlab="Year",
     main="Spline Slope (Effective Coefficient)")
```

Estimated change scores for early and late period

```{r}
# compute early and later change
# (i.e., before 2007)
print(paste('Change before 2007: Mean', mean(delta[year < 2007]),
            'SD', sd(delta[year < 2007]) / sqrt(length(delta[year < 2007]))))

print(paste('Change after 2007: Mean', mean(delta[year > 2007]),
            'SD', sd(delta[year >= 2007]) / sqrt(length(delta[year >= 2007]))))
```

```{r include = FALSE}
rm(ann_prod, ann_prod_plot, df, model_02, x,
   delta, dYear, N, year)
```

# Compute results for Figure 2a

First get thesaurus for references (i.e., contains info about misspelled refreences
and author names)

```{r}
thesaurus <- read.table('./data/thesaurus/most_cited_refs_thesaurus.tsv',
                        header = T, sep = '\t')
```

Now clean reference lists from duplicates

```{r}
# get custom function
source('./r_functions.R')

# clean reference lists
M_new <- clean_citations(M, 'CR', thesaurus)

M_short <- M_new %>% filter(TC > 10)
save(M_short, file = "./data/M_short_TC_10.RData")

M_short_2010 <- M_short %>% filter(PY > 2009)
save(M_short, file = "./data/M_short_PY_2010.RData")

```

Compute most cited references and format citations in accordance with
APA style

```{r}
# get number of citations per cited reference
cited <- as.data.frame(bibliometrix::citations(M_new, 'CR')$Cited)

# format names
require(dplyr)
require(tidyr)
require(stringr)
cited <- cited %>% 
  separate(CR, c('Author', 'Year', 'Source'), ',') %>%
  mutate(Name = paste(tolower(Author), 
                      ' (', str_trim(Year, 'left'), '), ',  
                      tolower(str_trim(Source, 'left')), 
                      sep = '')) %>%
  mutate(Name_Short = paste(tolower(Author),
                            ' (', str_trim(Year, 'left'), ')', 
                            sep = '')) %>%
  filter(Freq >= 50) %>%
  arrange(desc(Freq)) %>%
  mutate(Name = factor(Name, levels=rev(Name))) %>%
  select(Name, Year, Name_Short, Source, Freq) %>%
  mutate(Year = as.numeric(str_trim(Year, 'left')))

# APA style to match in text references
intext <- c('Schultz et al., 1997', 'Rescorla & Wagner, 1972',
  'Holroyd & Coles, 2002', 'Sutton & Barto, 1998', 
  'Friston, 2005', 'Pearce & Hall, 1980',
  'Miltner et al., 1997', 'Schultz & Dickinson, 2000',
  'Gehring & Willoughby, 2002', "O'Doherty et al., 2003", 
  'Montague et al., 1996', 'Rao & Ballard, 1999', 
  'Schultz, 1998', 'Behrens et al., 2007', 
  "O'Doherty et al., 2004", 'Friston, 2010',
  'Mackintosh, 1975', 'Gehring et al., 1993',
  'Botvinick et al., 2001', 'Delorme & Makeig, 2004')

# remove some references to avoid clutter in figures
cited$in_text <- intext
cited <- cited %>% 
  mutate(label = ifelse(in_text == "O'Doherty et al., 2003", '(1)',
                        ifelse(in_text == "O'Doherty et al., 2004", '(2)', 
                               ifelse(in_text == 'Rao & Ballard, 1999', '(3)', 
                                      ifelse(in_text == 'Schultz, 1998', '(4)', 
                                             in_text)))))

```

```{r}
cited_plot <- ggplot(cited, aes(x = Year, y = Freq, fill = Freq, label = label)) +
  geom_segment(aes(x = -Inf, y = 50, 
                   xend = -Inf, yend = 175),
               color = 'black', size = rel(0.5), linetype = 1) +
  geom_segment(aes(x = 1970, y = -Inf, xend = 2010, yend = -Inf),
               color = 'black', size = rel(0.5), linetype = 1) +
  geom_point(stat = 'identity', show.legend = F, size = 1.5,
             shape = 21) +
  geom_text_repel(seed = 46, size = 2.75, max.overlaps = Inf,
                  min.segment.length = unit(0.1, 'lines'),
                  segment.alpha = 0.5, segment.size = 0.30, force = 5.0,
                  nudge_x = -1.5, nudge_y = 1.0, box.padding = .25, point.padding = .5,
                  color = 'black') +
  scale_fill_viridis(option = 'B', begin = 0.1, end = 0.9) +
  coord_cartesian(ylim = c(35, 180), xlim = c(1970, 2012)) +
  scale_y_continuous(breaks = c(seq(50, 175, 25))) +
  scale_x_continuous(breaks = c(seq(1970, 2010, 5))) +
  theme(plot.margin = unit(c(10, 10, 10, 10), 'pt'),
        plot.title = element_text(color = 'black', size = 18, face = 'bold', 
                                  hjust = 0, family = 'Mukta'),
        axis.title.x = element_text(color = 'black', size = 16,
                                    margin = margin(t = 10), face = 'bold',
                                    family = 'Mukta'),
        axis.title.y = element_text(color = 'black', size = 16,
                                    margin = margin(r = 10), face = 'bold',
                                    family = 'Mukta'),
        axis.text.x = element_text(color = 'black', size = 14, family = 'Mukta'),
        axis.text.y = element_text(color = 'black', size = 14, family = 'Mukta'),
        panel.background = element_rect(fill = 'white'),
        panel.grid.major.x = element_line(size = 0.5, linetype = 'solid',
                                          colour = "gray98"),
        panel.grid.major.y = element_line(size = 0.5, linetype = 'solid',
                                          colour = "gray98"),
        panel.grid.minor = element_blank(),
        strip.text = element_blank(),
        strip.background = element_blank(),
        legend.position='right',
        legend.title = element_blank(),
        legend.text = element_text(size = 8),
        panel.spacing = unit(1, "lines")) +
  labs(y = 'Times cited', x = 'Publication year', 
       title = 'Most cited references'); cited_plot
ggsave(filename = './results/figures/Fig_2.tiff',
       plot = cited_plot, 
       width = 16.0, height = 11.0,  units = 'cm', dpi = 1200)
```

# Compute results for Supplemental Figure 1

```{r, fig.align = 'center', fig.width = 10, fig.height = 6}
# compute cumulative of citations
times_cited <- M_new %>%
        select(AU, TC) %>%
        arrange(desc(TC)) %>%
        mutate(cumsum_tc = cumsum(TC)) %>%
        mutate(index = 1:nrow(M_new))

# compute thresholds for cumulative distribution of citations in the data set
# at the "traditional" cut-off 80 %, and more conservative cut-offs 90 % and 95 %
threshold_80p <- max(times_cited$cumsum_tc) * 0.85
threshold_90p <- max(times_cited$cumsum_tc) * 0.90
threshold_95p <- max(times_cited$cumsum_tc) * 0.95

min_citations_80p <- min(times_cited[times_cited$cumsum_tc < threshold_80p, 'TC'])
min_citations_90p <- min(times_cited[times_cited$cumsum_tc < threshold_90p, 'TC'])
min_citations_95p <- min(times_cited[times_cited$cumsum_tc < threshold_95p, 'TC'])

# create plot showing citation thresholds
citation_threshold <- ggplot(data = times_cited, aes(x = index, y = TC)) +
        geom_segment(aes(x = -Inf, y = 0,
                         xend = -Inf, yend = 1600),
                     color = 'black', size = rel(0.5), linetype = 1) +
        geom_segment(aes(x = 1, y = -Inf, xend = 1400, yend = -Inf),
                     color = 'black', size = rel(0.5), linetype = 1) +
        scale_x_continuous(breaks = c(1, 200, 400, 600, 800, 1000, 1200, 1400))+
        scale_y_continuous(breaks = c(1, 200, 400, 600, 800, 1000, 1200, 1400, 1600))+
        geom_point(size = 0.8) +
        geom_line(aes(group = 1)) +
        geom_segment(aes(x = which.min(times_cited[times_cited$cumsum_tc < threshold_80p, 'TC']),
                         xend = which.min(times_cited[times_cited$cumsum_tc < threshold_80p, 'TC']),
                         y = 0, yend = 1500),
                         size = 0.5) +
        geom_segment(aes(x = which.min(times_cited[times_cited$cumsum_tc < threshold_95p, 'TC']),
                         xend = which.min(times_cited[times_cited$cumsum_tc < threshold_95p, 'TC']),
                         y = 0, yend = 1000),
                     size = 0.5) +
        annotate("label", x = which.min(times_cited[times_cited$cumsum_tc < threshold_80p, 'TC']) + 90,
                 y = 1525, label = paste0("80% threshold\n", min_citations_80p, " citations")) +
        annotate("label", x = which.min(times_cited[times_cited$cumsum_tc < threshold_95p, 'TC']) + 90,
                 y = 1025, label = paste0("95% threshold\n", min_citations_95p, " citations")) +
        theme(plot.margin = unit(c(10, 10, 10, 10), 'pt'),
              plot.title = element_text(color = 'black', size = 18, face = 'bold',
                                        hjust = 0, family = 'Mukta'),
              axis.title.x = element_text(color = 'black', size = 16,
                                          margin = margin(t = 10), face = 'bold',
                                          family = 'Mukta'),
              axis.title.y = element_text(color = 'black', size = 16,
                                          margin = margin(r = 10), face = 'bold',
                                          family = 'Mukta'),
              axis.text.x = element_text(color = 'black', size = 14, family = 'Mukta'),
              axis.text.y = element_text(color = 'black', size = 14, family = 'Mukta'),
              panel.background = element_rect(fill = 'white'),
              panel.grid.major.x = element_line(size = 0.5, linetype = 'solid',
                                                colour = "gray98"),
              panel.grid.major.y = element_line(size = 0.5, linetype = 'solid',
                                                colour = "gray98"),
              panel.grid.minor = element_blank(),
              strip.text = element_blank(),
              strip.background = element_blank(),
              legend.position='right',
              legend.title = element_blank(),
              legend.text = element_text(size = 8),
              panel.spacing = unit(1, "lines")) +
        labs(y = 'Times cited', x = 'Publication ID',
             title = 'Number of citations by publication'); citation_threshold
ggsave(filename = './results/figures/Fig_S1.tiff',
       plot = citation_threshold,
       width = 16.0, height = 11.0,  units = 'cm', dpi = 1200)

```



# Compute results for Figure 2b

First get thesaurus for keywords (i.e., contains info about misspelled keywords
and duplicates)

```{r}
# get custom function
source('./r_functions.R')

thesaurus_topics <- read.table('./data/thesaurus/trend_topics_thesaurus.tsv',
                               header = T)

```


Now clean the keyword lists

```{r }
M_new <- clean_citations(M, 'ID', thesaurus_topics)
```

Compute frequency of appearance measures for each keyword 
(only do this for timespan 2010 - 2020)

```{r} 
trends <- fieldByYear(M_new, field = "ID", timespan = c(2010, 2020),
                      min.freq = 5, n.items = 5, graph = F)

# adjust names to avoid clutter
trend_topics <- trends$df_graph %>% 
  arrange(desc(year_med)) %>%
  mutate(item = as.character(item)) %>%
  mutate(item = ifelse(item == 'episodic memory', 
                       "memory^(6)", item)) %>%
  mutate(item = ifelse(item == 'language comprehension', 
                       "language^(7)", item)) %>%
  mutate(item = ifelse(item == 'judgment', 
                         "(1)", item)) %>%
  mutate(item = ifelse(item == 'identification', 
                       "(2)", item)) %>%
  mutate(item = ifelse(item == 'attribution', 
                       "(3)", item)) %>%
  mutate(item = ifelse(item == 'awareness', 
                       "(4)", item)) %>%
  mutate(item = ifelse(item == 'communication', 
                       "(5)", item)) %>%
  mutate(item = str_replace(item, ' ', '~')) %>%
  ungroup() %>%
  mutate(item = factor(item),
         sfreq = (freq-min(freq))/(max(freq)-min(freq)))
```

Create Figure 3

```{r}
trends_plot <- ggplot(data = trend_topics, 
       aes(x = year_med, y = freq, label = item, fill = sfreq)) +
  geom_point(stat = 'identity', show.legend = F, size = 1.5,
             shape = 21) +
  geom_text_repel(seed = 46, size = 3.25, max.overlaps = Inf,
                  min.segment.length = unit(0.1, 'lines'),
                  segment.alpha = 0.5, segment.size = 0.30, force = 5.0,
                  box.padding = .35, point.padding = .5,
                  color = 'black', parse = T) +
  coord_cartesian(xlim = c(2008, 2022), ylim = c(-5, 110)) +
  scale_x_continuous(breaks = c(seq(2010, 2020, 5))) +
  scale_y_continuous(breaks = c(seq(5, 105, 25))) +
  geom_segment(aes(x = -Inf, y = 5, 
                   xend = -Inf, yend = 105),
               color = 'black', size = rel(0.5), linetype = 1) +
  geom_segment(aes(x = 2010, y = -Inf, xend = 2020, yend = -Inf),
               color = 'black', size = rel(0.5), linetype = 1) +
  scale_fill_viridis(option = 'B') +
  theme(plot.margin = unit(c(10, 10, 10, 10), 'pt'),
        plot.title = element_text(color = 'black', size = 20, face = 'bold', 
                                  hjust = 0, family = 'Mukta'),
        axis.title.x = element_text(color = 'black', size = 18,
                                    margin = margin(t = 10), face = 'bold',
                                    family = 'Mukta'),
        axis.title.y = element_text(color = 'black', size = 18,
                                    margin = margin(r = 10), face = 'bold',
                                    family = 'Mukta'),
        axis.text.x = element_text(color = 'black', size = 12, 
                                   family = 'Mukta'),
        axis.text.y = element_text(color = 'black', size = 12, 
                                   family = 'Mukta'),
        panel.background = element_rect(fill = 'white'),
        panel.grid.major.x = element_line(size = 0.5, linetype = 'solid',
                                          colour = "gray98"),
        panel.grid.major.y = element_line(size = 0.5, linetype = 'solid',
                                          colour = "gray98"),
        panel.grid.minor = element_blank(),
        strip.text = element_blank(),
        strip.background = element_blank(),
        legend.position='right',
        legend.title = element_blank(),
        legend.text = element_text(size = 8),
        panel.spacing = unit(1, "lines")) +
  labs(y = 'Occurrences', x = 'Year of appearance', 
       title = 'Trend topics'); trends_plot
ggsave(filename = './results/figures/Fig_3.tiff',
       plot = trends_plot, 
       width = 18, height = 15,  units = 'cm', dpi = 1200)
  
```
